---
title: "Exploring the ToothGrowth dataset"
author: "Spiros Baxevanakis"
date: "May 08, 2021"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
    highlight: tango
  html_document:
    toc: yes
    toc_depth: 3
    df_print: paged
  html_notebook:
    toc: yes
    number_sections: yes
editor_options:
  chunk_output_type: console
---

# Load and Explore Data

```{r, setup, results=FALSE, message=FALSE, warning=FALSE, tidy=TRUE, echo=F}
# load required libraries
library(ggplot2)
library(stringr)
library(tidyverse)
library(knitr)
```

One of the standard learning data sets included in R is the "ToothGrowth" data set. The tooth growth data set is the length of the odontoblasts (teeth) in each of 10 guinea pigs at three Vitamin C dosage levels (0.5, 1, and 2 mg) with two delivery methods (orange juice or ascorbic acid).

The file contains 60 observations of 3 variables:

-   len : Tooth length

-   supp : Supplement type (VC or OJ)

-   dose : Dose in milligrams

```{r}
library(datasets)
# convert df to tibble for tidyverse compatibility
tooth = tibble(ToothGrowth)
# convert dose column to factor
tooth$dose = parse_factor(as.character(tooth$dose))
```

```{r}
glimpse(tooth) # print the first few values
```

```{r}
summary(tooth) # calcuate some statistics for each variable
```

```{r}
count(tooth, supp, dose) # group by supp and dose, then count their observations
```

For every combination of supp and dose there are 10 observations (rows).

Now let's do some plots!

```{r}
# histogram by supp and dose
ggplot(data=tooth, aes(x=len)) +
    geom_density() +
    facet_grid(supp ~ dose) +
    labs(title = 'Density estimation for every combination of supp and dose',
         subtitle = 'Observe that as the dose increases the distribution
looks more skewed to the right. This indicates that larger
doses increase the tooth length more than smaller doses.')

# boxplot by supp and dose
ggplot(data=tooth, aes(y=len)) +
    geom_boxplot() +
    facet_grid(supp ~ dose) +
    labs(title = 'Boxplot for every combination of supp and dose',
         subtitle = 'Observe that as the doses get higher, the distribution of the
tooth length also contains larger values.')

# mean and var by supp
tooth %>% 
  group_by(supp) %>% 
  summarize(mean=mean(len),var=var(len))

# mean and var by dose
tooth %>% 
  group_by(dose) %>% 
  summarize(mean=mean(len),var=var(len))

# mean and var for every combination of supp and dose
tooth %>% 
    group_by(supp,dose) %>% 
    summarize(mean=mean(len),var=var(len))
```

The mean tooth length was larger when taking the OJ supplement with .5 and 1 doses. With the 2mg dose however there is no difference in the mean tooth length between supplement types. Although, at 2mg dose, the OJ supplement is 3 times less variable on tooth length than the VC supplement.

# Hypothesis & Permutation Tests

## Supplement Tests

Test the null hypothesis that the supplement type does not affect tooth length. $H_0$ is that the difference in means is 0 and $H_a$ that the difference in means is not 0.

```{r}
# select len column for every supplement type
toothoj = tooth %>% filter(supp == 'OJ') %>% select(len)
toothvc = tooth %>% filter(supp == 'VC') %>% select(len)
```

```{r}
t.test(toothoj, toothvc, alternative = 'two.sided')
```

The p value is very close to the rejection region and although normally we would fail to reject the $H_0$, we know from the exploratory data analysis we performed, that the $H_a$ is true. Since the p value is calculated in the direction of the $H_a$ we hypothesize that if we perform the same test using $H_a > 0$ instead of the original $H_a \neq 0$ the p value would be smaller allowing us to reject the $H_0$. This is because we observe from the data that the difference in means is greater than 0 (mean of OJ - mean of VC).

```{r}
t.test(toothoj, toothvc, alternative = 'greater')
```

Indeed, our hypothesis is correct. The p value is smaller when testing on Ha \> 0. Thus we can reject the Ho.

## Dose Tests

```{r}
# get all the observations that have doses 1 or 2
doses12 = tooth %>% filter(dose %in% c(1,2)) %>% select(len, dose)
y = doses12$len # get only len variable
dose = doses12$dose # get only dose variable 
# function to calcuate differnece in means
testStat = function(y,d) mean(y[d==2]) - mean(y[d==1])
observedStat = testStat(y,dose)
# sample dose labels and calculate testStat 10000 times.
permutations = sapply(1:10000, function(i) testStat(y, sample(dose)))
observedStat
```

```{r}
# number of permutations that are more extreme than the observed stat
mean(permutations > observedStat)
```

No permutation is as or more extreme than our observed statistic (except the observed statistic itself)! We can safely conclude that a dose of 2 is associated with a larger tooth length. This is also observed in the box and density plots of the previous sections.

We will repeat the same procedure between .5 and 1 doses.

```{r}
doses51 = tooth %>% filter(dose %in% c(.5,1)) %>% select(len, dose)
y = doses51$len
dose = doses51$dose
testStat = function(y,d) mean(y[d==1]) - mean(y[d==.5])
observedStat = testStat(y,dose)
permutations = sapply(1:10000, function(i) testStat(y, sample(dose)))
observedStat
```

```{r}
mean(permutations > observedStat)
```

We reach the same conclusion as with the comparison between 2 and 1 doses.

# Assumptions

In order to draw the conclusions in the hypothesis test section, we assume that the data is not skewed and follows a normal distribution. This is a requirement of the Student's t-test and t-confidence interval.
